Community-conditioned language models for linguistic variation

### 2021-10-12 paper/paper1.tex
various small changes which are essentially procrastination

\title{Community-Conditioned Language Models}

@@

\begin{document}
\maketitle
\begin{abstract}
  Community-level linguistic variation is a core concept in sociolinguistics,
  but more tools are needed to study variation from a computational perspective.
  In this paper, we use conditional language models to investigate linguistic variation
  across 510 online communities.
  We investigate the community representations learned by the conditional language model
  and find that they are highly correlated with a social network-based 
  representation that doesn't use any linguistic input.
\end{abstract}

@@

\section{Introduction}
Linguistic communication requires that speakers share
knowledge of certain linguistic conventions, such as syntactic
structure, word meanings, and accepted patterns of interaction.
Speakers assume that these conventions are \emph{common ground} among
their interlocutors, based on joint membership in a community
\cite{Stalnaker2002, Clark1996}.  Such a community may be of any size,
from the very small, like members of a particular friend group, to the
very large, like speakers of English.  Because it is the source of
linguistic convention, the \emph{speech community} is an important
concept in the study of linguistic variation \citep{Gumperz1972}.

% Not only is linguistic variation an object of inquiry in its own
% right, but it is also important to consider it in the development of
% robust, equitable language technology.  Previous work has shown that
% by taking demographic information into account, models can achieve
% better performance on various standard NLP tasks
% \citep{Hovy2015,Yang2017}. \jp{Move this to related work so we can get to the point quicker.}
Most of the previous computational work on linguistic variation 
has considered variation at the level of macro-social categories, such as gender
\citep{Burger2011,Ciot2013,Bamman2014}, age \citep{Nguyen2013}, and
geographic location \citep{Eisenstein2010,Bamman2014a}.
Some recent exceptions are \citet{DelTredici2017} who use a modified
skip-gram model to show that lexical semantic variation occurs across 
different communities organised around the same topic and 
\citet{Lucy2021}, who use contextualised word representations to
investigate community-level variation in word sense distributions.

In the present work, we investigate linguistic variation
across different online communities. In contrast to previous work, we
investigate variation across speech communities (rather than
macro-social categories).

@@

For this purpose, we define (\cref{sec:cclm}) 
various Community Conditioned Language Models 
(CCLM for short).
% , and model the degree to which a
% given message $m$ is acceptable in a given community $c$, by estimating the probability
% \(P(m \mid c)\).

We want to find out how, and how much, CCLMs are attuned to
In practice, we test whether the linguistic
models are correlated with a social network model of communities (based only on
co-occurrence of users) (\cref{sec:embedding-analysis})

@@

\section{Community-conditioned language models}\label{sec:cclm}

We experiment with two kinds of model architecture: a simple
unidirectional LSTM \citep{Hochreiter1997} and a Transformer
\citep{Vaswani2017}.  In either case, the model is organised as a
standard $n$-layer neural sequence encoder (we set $n=3$ for all
models). As is convention, a sequence of word tokens is first converted to 
vectors using a trainable embedding layer over a pre-determined vocabulary.
At the other end, word tokens are predicted using a softmax projection 
layer. What we have described so far does not take community into account
and as such we call them \emph{uncodnitioned models}, but the same
encoder architecture also forms the core of our conditioned models.

To condition the model , we add a \emph{community embedding} parameter.
This parameter is concatenated with the hidden layer of the
sequence encoder, at some layer $l_c \leq n$, and passed through a
linear layer which projects the resulting vector back to the original
hidden layer size.  For $l_c = n$, the output of this linear layer is
passed directly to the softmax function, just as the final hidden
layer of the sequence encoder is in other models.  For $l_c=0$, the
community embedding is concatenated with the token embedding.  For
this reason, we set the hidden size of the sequence encoder and the
size of the token embedding to be equal for all models.

@@

\subsection{Results and analysis}

In this section, we report the performance of the conditioned and un-conditioned
models on the test set.
First, we define two performance metrics, perplexity and information gain.
In the following, we use $M$ to refer to messages in the
combined test set, and $M_j$ for the partition of the test set originating from 
community $c_j$.

@@

For a given model let $H(m)$ be the model's cross-entropy loss,
averaged over tokens in $m$.
We define the perplexity on a set of messages, $M$,
to be the exponential of the model's average cross-entropy loss:
\[\Ppl_M = e^{\avg_{m\in M} H(m)}\]

@@

We also consider the average information gain per token of the CCLM over its baseline
un-conditioned counterpart, with the same sequence encoder architecture.
For a given message, information gain is defined as the difference
between the cross-entropy of the unconditioned model and the conditioned model:
\[H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m)\]
For a set of messages, $M$, we consider the average information gain
in exponential space (as a ratio of perplexities):
\[\IG_M = \frac{e^{\avg_{m \in M}(H_{\mathrm{LM}}(m))}}{e^{\avg_{m \in M}(H_{\mathrm{CCLM}}(m))}}\]
\[\IG_M = {e^{\avg_{m \in M}(H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m))}}\]
%
Unsurprisingly, the conditioned models mostly have lower perplexity 
than their respective unconditioned baseline models, 
meaning information gain above one (\cref{tab:model-results}).
Overall, the LSTM models perform better than the transformers, 
although the best transformer has higher information gain than the best LSTM.

@@

We also consider performance stratified by community; that is,
$\Ppl_{M_j}$ and $\IG_{M_j}$, 
where $M_j$ is the set of messages originating from community $c_j$ 
(\cref{fig:comm-stratified-box}).
We observe a lot of variation in baseline perplexity
across communities, with $\Ppl_{M_j}$ ranging from \num{3.67} to
\num{93.58} for the best conditional LSTM model 
(see supplementary materials for perplexity by community). %TODO keep? 
While the absolute performance ($\Ppl_{M_j}$) of the LSTM models is better,
the best transformer models have somewhat higher information gain
than their LSTM counterparts.

@@

The conditioned models also perform differently across different communities---%
even among the best models, some communities have $\IG_{M_j} < 1$,
meaning that the CCLM performs worse than the unconditioned baseline for 
messages from that community.
For other communities $\IG_{M_j}$ is much higher, meaning that the CCLM performs better  
(\cref{fig:comm-stratified-box}).\footnote{
  Some of the communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English, 
  but surprisingly, not the three most extreme
  outliers. There are \subreddit{counting}, \subreddit{friendsafari}, and \subreddit{Fireteams}, 
  the later two of which are places where people coordinate to play video games
  together. The coordination has highly structured and regular formats, which 
  are presumably conventional to these communities.}
%
%The variation in information gain across communities is not 
%related to perplexity, however (\cref{fig:ppl-info-gain}).

We observe empirically that across all the models we
tested, communities where conditioning has the least effect tend to be
more organised around more general interest topics, such as 
\subreddit{relationships} and \subreddit{advice}, where the subject
matter is relevant to a broad range of people.  Conditioning the model
on community has the most benefit for narrower special-interest
subreddits, such as those organised around a certain videogame, sports
team, or subculture.  This makes sense intuitively, since communities
with more niche topics would tend to have more specialised vocabulary
and other linguistic patterns.

@@

In this section we investigate whether (and how much) the community
embeddings obtained from the CCLMs correlate with the community-member
relationship on Reddit.

To this end, we compare the CCLM-learned community
embedding with the community embedding created by \citet{Kumar2018},
who generated them using using a negative-sampling optimization
algorithm and the message-community co-occurrence matrix as ground
truth, using data which spans January 2014 to April 2017.
We refer the reader
to \citet{Kumar2018}
for details, but the important point is that no linguistic
information is used to create these embeddings: they only reflect the social
relationship between communities, via 
the user-community membership relation. In contrast, CCLM community embeddings depend in no
way on which user is the author of any given message: we only use the
contents of messages, not authorship data.
%

\subsection{Comparing embeddings: anecdotal evidence}
\label{sec:storytelling}
TODO: BILL

@@

\[\C = {(x = L_i · L_j, y = S_i · S_j) \text {~for~} i,j ∈ [1,512]\}\]

@@

Now, to test if \(d(L,S)\) corresponds to a significant correlation,
it suffices to check if its value is significantly larger than it
value for random linguistic embeddings \(L\). Even though this is the
distribution of \(d(L,S)\) for random embeddings is difficult to
compute analytically, one can instead evaluate it using a Monte Carlo
method.

>>>>

\title{Community-conditioned language models for linguistic variation}

@@

\begin{document}
\maketitle
\begin{abstract}
  Community-level linguistic variation is a core concept in sociolinguistics,
  but more tools are needed to study variation from a computational perspective.
  In this paper, we use conditional language models to investigate linguistic variation
  across 510 online communities.
  We probe the community representations learned by the conditional language model
  and find that they are highly correlated with a social network-based 
  representation that doesn't use any linguistic input.
\end{abstract}

@@

\section{Introduction}
Linguistic communication requires that speakers share
certain linguistic conventions, such as syntactic
structure, word meanings, and patterns of interaction.
Speakers assume that these conventions are \emph{common ground} among
their interlocutors, based on joint membership in a community
\cite{Stalnaker2002, Clark1996}.  Such a community may be of any size,
from the very small, like members of a particular friend group, to the
very large, like speakers of English.  Because it is the source of
linguistic convention, the \emph{speech community} \citep{Gumperz1972}
is an important concept in the study of linguistic variation.

% Not only is linguistic variation an object of inquiry in its own
% right, but it is also important to consider it in the development of
% robust, equitable language technology.  Previous work has shown that
% by taking demographic information into account, models can achieve
% better performance on various standard NLP tasks
% \citep{Hovy2015,Yang2017}. \jp{Move this to related work so we can get to the point quicker.}
Most previous computational work on linguistic variation 
has considered variation at the level of macro-social categories, such as gender
\citep{Burger2011,Ciot2013,Bamman2014}, age \citep{Nguyen2013}, and
geographic location \citep{Eisenstein2010,Bamman2014a}.
Recent exceptions are \citet{DelTredici2017} who use a modified
skip-gram model to show that lexical semantic variation occurs even across 
different communities organised around the same topic and 
\citet{Lucy2021}, who use contextualised word representations to
investigate community-level variation in word sense distributions.

In the present work, we investigate linguistic variation
in online communities on the social media website Reddit. 
In contrast to previous work, we investigate variation across 
speech communities, rather than macro-social categories.

@@

For this purpose, we introduce (\cref{sec:cclm}) various 
Community-Conditioned Language Models (CCLMs for short),
which learnan embedding of the communites they are conditioned on 
as a side effect of their language modeling objective.
We test which architectures best make use of the community information and 
how and in what ways the resulting embeddings are
correlated with a social network-based representation of communities(\cref{sec:embedding-analysis}).
% , and model the degree to which a
% given message $m$ is acceptable in a given community $c$, by estimating the probability
% \(P(m \mid c)\).

@@

\section{Community-conditioned language models (CCLMs)}\label{sec:cclm}

We experiment with two kinds of model architecture: a simple
unidirectional LSTM \citep{Hochreiter1997} and a Transformer
\citep{Vaswani2017}.  In either case, the model is organised as a
standard $3$-layer neural sequence encoder, where the input for the
$t$th timestep of the $n+1$st layer is the $t$th hidden state of the 
$n$th layer.
As usual, the input to the first layer, is a sequence of tokens,
encoded with a trainable embedding layer over a pre-determined vocabulary.
At the other end, word tokens are predicted using a softmax projection 
layer. What we have described so far does not take community into account
and as such we call them \emph{uncodnitioned models}, but the same
encoder architecture also forms the core of our conditioned models.

In the conditional models, we add a \emph{community embedding} parameter.
This parameter is concatenated (at each time step) with the hidden layer of the
sequence encoder, at some layer $l_c \leq n$, and passed through a
linear layer which projects the resulting vector back to the original
hidden layer size.  For $l_c = n$, the output of this linear layer is
passed directly to the softmax function, just as the final hidden
layer of the sequence encoder is in other models.  For $l_c=0$, the
community embedding is concatenated with the token embedding.  For
this reason, we set the hidden size of the sequence encoder and the
size of the token embedding to be equal for all models.

@@

\section{CCLM Performance}

In this section, we report the performance of the conditioned and un-conditioned
models on the held out test set.
First, we define two performance metrics: perplexity and information gain.
In the following, we use $M$ to refer to messages in the
combined test set, and $M_j$ for the partition of the test set originating from 
community $c_j$.

@@

For a given model, let $H(m)$ be the model's cross-entropy loss,
averaged over tokens in $m$.
We define the perplexity on a set of messages, $M$,
to be the exponential of the model's average cross-entropy loss:
\[\Ppl_M = e^{\avg_{m\in M} H(m)}\]

@@

We also consider the average information gain per token of the CCLM over its baseline
un-conditioned counterpart, with the same sequence encoder architecture.
For a given message, information gain is defined as the difference
between the cross-entropy of the unconditioned model and the conditioned model:
\[H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m)\]
For a set of messages, $M$, we consider the average information gain
in exponential space (as a ratio of perplexities):
\[\IG_M = \frac{e^{\avg_{m \in M}(H_{\mathrm{LM}}(m))}}{e^{\avg_{m \in M}(H_{\mathrm{CCLM}}(m))}}\]
\[\IG_M = {e^{\avg_{m \in M}(H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m))}}\]
%
Unsurprisingly, the conditioned models mostly have lower perplexity 
than their respective unconditioned baseline models, 
meaning information gain above one (\cref{tab:model-results}).
While the absolute performance ($\Ppl_{M_j}$) of the LSTM models is better,
the best transformer models have somewhat higher information gain
than their LSTM counterparts.

@@

We also consider performance stratified by community; that is,
$\Ppl_{M_j}$ and $\IG_{M_j}$, 
where $M_j$ is the set of messages originating from community $c_j$ 
(\cref{fig:comm-stratified-box}).
We observe a lot of variation in baseline perplexity
across communities, with $\Ppl_{M_j}$ ranging from \num{3.67} to
\num{93.58} for the best conditional LSTM model 
(\cref{fig:comm-stratified-box}; also see supplementary materials for perplexity by community). %TODO keep? 
The conditioned models also perform differently across different communities---%
even among the best models, some communities have $\IG_{M_j} < 1$,
meaning that the CCLM performs worse than the unconditioned baseline for 
messages from that community.
For other communities $\IG_{M_j}$ is much higher, meaning that the CCLM performs better  
(\cref{fig:comm-stratified-box}).\footnote{
  Some of the communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English, 
  but surprisingly, not the three most extreme
  outliers. There are \subreddit{counting}, \subreddit{friendsafari}, and \subreddit{Fireteams}, 
  the later two of which are places where people coordinate to play video games
  together. The messages in these communites adhere to highly regular formats, 
  which are presumably conventional to the communities.}
%
%The variation in information gain across communities is not 
%related to perplexity, however (\cref{fig:ppl-info-gain}).

@@

We observe empirically that across all the models we
tested, communities where conditioning has the least effect tend to be
organised around more general interest topics, such as 
\subreddit{relationships} and \subreddit{advice}, where the subject
matter is relevant to a broad range of people.  Conditioning the model
on community has the most benefit for narrower special-interest
subreddits, such as those organised around a certain videogame, sports
team, or subculture.  This makes sense intuitively, since communities
with more niche topics would tend to have more specialised vocabulary
and other linguistic patterns.

@@

In this section we investigate whether (and how much) community
embeddings obtained from the CCLMs correlate with the community-member
relationship on Reddit.

To this end, we compare the CCLM-learned community
embedding with the community embedding created by \citet{Kumar2018},
which were generated using using a negative-sampling optimization
algorithm, with the message author-community co-occurrence matrix as ground
truth, using data from January 2014 to April 2017.
We refer the reader
to \citet{Kumar2018}
for details, but the important point is that no linguistic
information is used to create these embeddings: they only reflect the social
relationship between communities, via 
the user-community membership relation. In contrast, CCLM community embeddings depend in no
way on which user is the author of any given message: we only use the
contents of messages, not authorship data.
%

\subsection{Comparing embeddings: anecdotal evidence}
\label{sec:storytelling}

To get a handle on the relationship between the linguistic and social embeddigs

@@

\[C = \{(x = L_i · L_j, y = S_i · S_j) \text {~for~} i,j ∈ [1,512]\}\] % BN: changed \C to C. Definition for \C may be missing

@@

Now, to test if \(d(L,S)\) corresponds to a significant correlation,
it suffices to check if its value is significantly larger than the same
value for random linguistic embeddings \(L\). Even though this is the
distribution of \(d(L,S)\) for random embeddings is difficult to
compute analytically, one can instead evaluate it using a Monte Carlo
method.
